---
title: "rerandR: Re-randomization approach"
author: "Kevin W. McConeghy"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rerandR: 02-rerand}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: mybib.bib
---

Welcome to rerandR! This program was written as a companion to academic work performed by researchers at Brown University to conduct re-randomization for cluster-randomized nursing home trials. 

```{r, echo = FALSE, message = FALSE}
library(tidyverse)
library(ggplot2)
library(rerandR)
```

## Introduction

In this vignette we will explain how to use covariate balance as criteria for re-randomization and introduce the companion functions. The basic procedure outlined by @MorganKL2015 is as follows:  

  1) Select units for the comparison of treatments, and collect covariate data on all units.  
  
  2) Define an explicit criterion for covariate balance.  
  
  3) Randomize units to treatment groups.  
  
  4) Check covariate balance and return to Step 3 if the allocation is unacceptable according to the criterion specified in Step 2; continue until the balance is acceptable.  
  
  5) Conduct the experiment.  
  
  6) Perform inference (using a randomization test that follows exactly steps 2-4).  

### Example dataset  
The example dataset used in this package is freely available and comes from a clinical AIDs trial conducted in the 1990s.[@HammerSM1996] See: ?rerandR::ACTG175

#### Load dataset  
```{r }
df <- rerandR::ACTG175
```

## Measure of covariate balance  

We will use Mahalanobis distance with sample size constant to estimate covariate balance, acceptance probability and identify an acceptable randomization.[@MorganKL2015]  

$$ M \equiv \frac{n_tn_c}n (\bar{X_t} - \bar{X_c}) ' cov(X)^{-1} (\bar{X_t} - \bar{X_c}) $$
Where n is the sample size, t treated, c controls, and X represents the covariate means. 

### Trial Covariate Balance  
The ACT trial was a single randomized trial, here are the observed mean differences +/- standard deviation in that single randomization.  
```{r, eval=F}
df_grp_means <- df %>%
  select(arms, age, race, gender, symptom, wtkg, hemo, msm, 
         drugs, karnof, oprior) %>%
  group_by(arms) %>%
  summarize_all(., mean) %>%
  t(.) 
colnames(df_grp_means) <- c('zido', 'combo')

df_grp_sd <- df %>%
  select(arms, age, race, gender, symptom, wtkg, hemo, msm, 
         drugs, karnof, oprior) %>%
  summarize_all(sd) %>%
  t(.)
  
t <- df_grp_means[, 1] - df_grp_means[, 2] %>%
  bind_cols(mean = ., sd = df_grp_sd) 

t %>% round(., 4)
```

```{r }
df_t <- df[df$arms==1, ] %>%
  select(age, race, gender, symptom, wtkg, hemo, msm, 
         drugs, karnof, oprior)

df_c <- df[df$arms==0, ] %>%
  select(age, race, gender, symptom, wtkg, hemo, msm, 
         drugs, karnof, oprior) 

cat('The M-distance for this trial, estimated using patient characteristics is:  \n', mdist(df_t, df_c))
```

But is this a good or bad randomization?  How to get some context:

#### Distribution of M  
Assuming normality of covariate means, the M statistic is Chi-square distributed with k degrees of freedom corresponding to number of covariates. 
The alternative is to conduct a randomization test, compute M-distance for each randomization and use the empirical distribution of M to identify an acceptable cut-off.  

### M-distance - Chi-sq.  
Using guidance from Morgan & Rubin paper, an acceptance probability of 0.001 is a reasonable starting point.[@MorganKL2015] This is saying that we would like to find a M-distance cut-off which represents the lower 0.1% of the distribution of M (i.e. the most balanced randomization draws).  

```{r }
k <- length(df_t) #number of covariates, degrees of freedom
Pa <- 0.001

chi_val <- qchisq(0.001, k)

cat('M-distance cut-off using 10 covariates is:\n ', round(chi_val,2))
```

### M-distance - empirical distribution  
Here we perform a Permutation with 10,000 simulations, where randomly assign individuals to treatment group. We then compute the M-statistic and store the value.  

```{r }
permute_m <- function(df) {
  
  df_2 <- df %>%
    mutate(arms = rbinom(nrow(df), 1, 0.5)) #fair coin assigment
  
  df_t <- df_2[df_2$arms==1, ] %>%
    select(age, race, gender, symptom, wtkg, hemo, msm, 
         drugs, karnof, oprior)

  df_c <- df_2[df_2$arms==0, ] %>%
    select(age, race, gender, symptom, wtkg, hemo, msm, 
         drugs, karnof, oprior)
  
  return(mdist(df_t, df_c))
}

sims <- replicate(n=10000, permute_m(df))
```

```{r }
label <-  paste0('Pa, 0.001 = ', round(quantile(sims, 0.001), 4))
sims %>%
  tibble(x=.) %>%
  ggplot(., aes(x=x)) +
    geom_histogram(color = 'darkblue', fill = 'lightblue') +
    geom_vline(aes(xintercept=quantile(sims, 0.001)),
            color="blue", linetype="dashed", size=1) +
    annotate("text", x = quantile(sims, 0.975), y = 500, label=label) 
```
The empirical distribution suggests an acceptance probability of 0.001 is seen at M values of 1.36 or lower.  

At this point we have completed steps 1-3, we have our entire cohort, a list of covariates, and a defined criterion for accepting a randomization.  
Now we must conduct the re-randomization procedure...

### Re-randomization  

# Contacting authors  
The primary author of the package was Kevin W. McConeghy. [See here](https://github.com/kmcconeghy/)

# References  
