---
title: "jumble: Random assignment"
author: "Kevin W. McConeghy"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{jumble: 03-assignment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: mybib.bib
---

Welcome to jumble! This program was written as a companion to academic work performed by researchers at Brown University to conduct re-randomization for cluster-randomized nursing home trials. 

```{r, echo = FALSE, message = FALSE}
library(tidyverse)
library(ggplot2)
library(jumble)
```

## Introduction

In this vignette we demonstrate some functions used for random assignment, following best practices outlined by other R packages like randomizeR, randomizr etc.  

  1) Randomization must be transparent
  2) Randomization must be reproducibile
  3) Randomization must be random!

### Example dataset  
The example dataset used in this package is freely available and comes from a clinical AIDs trial conducted in the 1990s.[@HammerSM1996] See: ?jumble::ACTG175

#### Load dataset  
```{r }
df <- jumble::ACTG175
```

## Random assignment to K-groups  
Assigning treatment is easy if you want equal probability of assignment to one or more groups, without concern to final sample size in each arm, covariate balance etc.  

```{r }
assign <- rnd_assign(df$pidnum, 2, as.integer(as.Date('2019-06-26')))

head(assign)
```
Truly random assignment!  

The seed is stored as a attribute  
```{r }
seed_iter <- attr(assign, 'seed')
```
```{r, echo=F}
paste0('Randomization performed: ', Sys.Date(), ' Seed: ', seed_iter)
```

Don't lose that information!  As long as you know the seed you can reproduce the randomization:  
```{r }
assign <- rnd_assign(df$pidnum, 2, seed_iter)

head(assign)
```

### Simulation to demonstrate equal probability  
Perform 20,000 randomizations  
```{r }
pidnum <- df$pidnum  

n=20000L

df_assign <- lapply(1:n, function(x) rnd_assign(pidnum, 2, 18073L+x))

assigns <- bind_cols(df_assign[[1]][, 1], 
                     lapply(df_assign, function(x) x[, 2]))

df_assign_2 <- gather(assigns, key = 'trial', value = 'group', 
                      starts_with('group')) %>%
  group_by(id) %>%
  summarize(`No. of trials` = n,
            `No. group 'a'` = sum(group == 'a'),
            `No. group 'b'` = sum(group == 'b'),
            `Prob. group = 'a'` = `No. group 'a'` / n(),
            `Prob. group = 'b'` = `No. group 'b'` / n())

head(df_assign_2)
```
First glance shows ~ equal likelihood of group assignment in 20,000 trials. 
### Assignment tests  
```{r, echo=F}
ggplot(df_assign_2, aes(x = `Prob. group = 'a'`,
                        y = `Prob. group = 'b'`)) +
  geom_violin( colour = 'blue') +
  scale_x_continuous(limits = c(0.47, 0.53)) +
  scale_y_continuous(limits = c(0.47, 0.53)) +
  labs(title="Probability of treatment assignment", 
       subtitle="K=2; completely random assignment; 20,000 trials",
       x="Pr | Group A",
       y="Pr | Group B")

```

For any given individual, across 20,000 trials no assignment exceeds an imbalance of ~2-3%.  

```{r, echo=F}
cat("Probability of assignment to group 'a' in 20,000 trials \n")
summary(df_assign_2$`Prob. group = 'a'`)
cat("Probability of assignment to group 'b' in 20,000 trials \n")
summary(df_assign_2$`Prob. group = 'b'`)
```

However, in each trial the number of individuals in group A vs. B was random and could be imbalanced.  

```{r }
df_trials <- gather(assigns, key = 'trial', value = 'group', 
                      starts_with('group')) %>%
  group_by(trial) %>%
  summarize(`No. group 'a'` = sum(group == 'a'),
            `No. group 'b'` = sum(group == 'b'))
```

```{r, echo=F}
cat("No. assigned to group 'a' in 20,000 trials \n")
summary(df_trials$`No. group 'a'`)
cat("No assigned to group 'b' in 20,000 trials \n")
summary(df_trials$`No. group 'b'`)
```

See how the number assigned varies widely?  

what if you want to guarantee an equal number of individuals are in each group?  

### Random assignment to K-groups, equal sample size  
We will repeat the exercise, dropping one person to make an odd-numbered cohort.  
```{r }
pidnum <- df$pidnum 
pidnum <- pidnum[2:length(pidnum)] # drop 1 

n=10000

df_assign <- lapply(1:n, function(x) rnd_allot(pidnum, 18073L+x))

assigns <- bind_cols(df_assign[[1]][, 1], 
                     lapply(df_assign, function(x) x[, 2]))

df_assign_2 <- gather(assigns, key = 'trial', value = 'group', 
                      starts_with('group')) %>%
  group_by(id) %>%
  summarize(`No. of trials` = n,
            `No. group 'a'` = sum(group == 'a'),
            `No. group 'b'` = sum(group == 'b'),
            `Prob. group = 'a'` = `No. group 'a'` / n(),
            `Prob. group = 'b'` = `No. group 'b'` / n())

head(df_assign_2)
```

### Assignment tests  
```{r, echo=F}
ggplot(df_assign_2, aes(x = `Prob. group = 'a'`,
                        y = `Prob. group = 'b'`)) +
  geom_violin( colour = 'blue') +
  scale_x_continuous(limits = c(0.47, 0.53)) +
  scale_y_continuous(limits = c(0.47, 0.53)) +
  labs(title="Probability of treatment assignment", 
       subtitle="K=2; completely random assignment; 20,000 trials",
       x="Pr | Group A",
       y="Pr | Group B")

```

Random assignment looks well-balanced between groups.  

```{r, echo=F}
cat("Probability of assignment to group 'a' in 20,000 trials \n")
summary(df_assign_2$`Prob. group = 'a'`)
cat("Probability of assignment to group 'b' in 20,000 trials \n")
summary(df_assign_2$`Prob. group = 'b'`)
```

```{r }
df_trials <- gather(assigns, key = 'trial', value = 'group', 
                      starts_with('group')) %>%
  group_by(trial) %>%
  summarize(`No. group 'a'` = sum(group == 'a'),
            `No. group 'b'` = sum(group == 'b'))
```

What about the size of the groups?  
```{r, echo=F}
cat("No. assigned to group 'a' in 20,000 trials \n")
summary(df_trials$`No. group 'a'`)
cat("No assigned to group 'b' in 20,000 trials \n")
summary(df_trials$`No. group 'b'`)
```
Equal size!  The number now randomly goes between 526 or 527 for either the 'a' or 'b' group. The imbalance is selected with equal probability.  

# Contacting authors  
The primary author of the package was Kevin W. McConeghy. [See here](https://github.com/kmcconeghy/)

# References  
