---
title: "jumble: Stratified Randomization"
author: "Kevin W. McConeghy"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{jumble: 04-stratification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: mybib.bib
---

Welcome to jumble! This program was written as a companion to academic work performed by researchers at Brown University to perform different randomization strategies with cluster-randomized nursing home trials. 

```{r, echo = FALSE, warning=F, message = FALSE}
library(tidyverse)
library(ggplot2)
library(jumble)
```

## Introduction  
In this vignette we go over how to do a stratified randomization, both with a single category and using a distance measure.  

### Example dataset  
The example dataset used in this package is freely available and comes from a clinical HIV therapy trial conducted in the 1990s.[@HammerSM1996] See: ?jumble::ACTG175

*Note* An important limitation of most clinical trials is that enrollment is graduated with individual patients consented and enrolled one at a time. However, cluster randomized trials can often identify all potential clusters before randomization occurs which allows more finessed randomization techniques.  

#### Load dataset  
```{r }
df <- jumble::ACTG175
```

## Single covariate stratification  
A common method for stratification is to take 1-2 covariates which are considered to be very important clinically for the outcome. Often a cluster is used (i.e. randomized within facility, hospital, village). In a cluster randomized trial, a variable which is correlated with the cluster may be more likely to be imbalanced so if you were particularly concerned about one thing that would be a potential.  

### Identify a strong predictor of the primary outcome  
Allow the trial design is a little more complicated, we will simply look at the `cens` event (A CD4 T cell drop of 50 or more, AIDs defining event, or death).  
```{r }
glm(cens ~ age + hemo + msm + drugs + oprior + z30 + 
      race + gender + str2 + symptom, data = df) %>%
  summary(.)
```
We identify two strong predictors, `drugs` which is history of prior IV drug abuse, and `symptom` which is an indictor for a symptomatic HIV infection on presentation (essentially a progressed infection).  

In general if you have no prior knowledge, you should select strata which are highly predictive of outcome.  

For an example, lets examine symptomatic presenters.  So maybe we want to conduct a subsequent trial where we group individuals by `symptom` then randomize within those two groups.  

### Random assignment within `symptom`  
```{r }
df_str <- rnd_str(df, symptom)
```

```{r, echo=F}
table(df_str$symptom, df_str$assign)
```

# Contacting authors  
The primary author of the package was Kevin W. McConeghy. [See here](https://github.com/kmcconeghy/)

# References  
