---
title: "jumble: Simulation Study Code"
author: "Kevin W. McConeghy"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{jumble: 05-simstudy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: mybib.bib
---

# Setup
```{r setup, eval=FALSE}
#clear
  knitr::opts_chunk$set(warning = F, message = F)
  source('P:\\flunm\\K1M\\flublok_pretrial\\flublok_pretrial_cfg.R')
```

## Load Cohort File 
```{r, eval=F}
filepath <- paste0(DataFiles, FileNames[[1, '1']])

df_trial <- readRDS(filepath) %>%
  mutate(female = if_else(mds_gender=='Female', 1L, 0L),
         pcthmo = pcthmo / 100) 
rw_title <- 'Base Cohort File'

des_df(df_trial, rw_title, c('bene_id_18900', 'fac_id'))
```

7 Methods 

1. Simple Random
2. 2-strata randomization
3. Pair-matched design
4. K-means cluster, a prior covariates
5. K-means cluster, PCA
6. Re-randomization

# Simulation Set-up 

## Parameters  

```{r, eval=F}
if (F) {
  n_rndms <- 100L
  n_re_rndms <- 100L
  n_permutes <- 10L
  k_clust <- 30L
  k_starts <- 100L
  k_iters <- 100L
} 

if (T) {
  n_rndms <- 10000L
  n_re_rndms <- 1000L
  k_clust <- 30L
  n_permutes <- 250L
  k_starts <- 200000L
  k_iters <- 100L
}

pr_reject <- 0.02  

st_seed <- as.integer(ymd('2019-12-26'))
set.seed(st_seed)

rndm_methods <- c('Simple Randomizations', 
             '2-Strata randomization',
             'Pair-matched design', 
             'K-means, covariates',
             'K-means, PCA',
             'Re-randomization')

#unique facility list
df_fac <- df_trial %>%
    pull(fac_id) %>%
    unique(.) 

cat('Starting Seed: ', st_seed, '\n')
cat('No. of random simulations performed: ', n_rndms, '\n')
cat('No. of permutations performed: ', n_permutes, '\n')
```

# Variable Spefications  

## Select key covariates  

```{r, eval=F }
df_cov <- df_trial %>%
  select(starts_with('fac'), "adefscore", 'dc_hosp_any',
  "facpoor","totbeds","alzunit","anyunit" ,"paymcaid", "paymcare", "payother", "multifac", "profit" ,"hospbase","restrain" , "acuindex2",
  "anymdex", "rn2nrs", "dchrppd", "rnhrppd", "lpnhrppd", "cnahrppd", "adm_bed", "agg_cmi_2011p", "agglocare_2011p", "agg_hosp", 
  "agg_comm", "aggadl_2011p", "agglowcfs", "aggmidcfs", "agghighcfs", "agg_female",  "aggblack_2011p", "agghisp_2011p", "aggwhite_2011p",
  "pcthmo", "agg_u65", "nresid", "avgage", "avgadl_2011p", "avgrugcmi_2011p", "pctlocare_2011p", "pctfem", 
  "pctblack_2011p", "pcthisp_2011p", "pctwhite_2011p", "pctunder65", "pctlowcfs", "pctmidcfs", "pcthighcfs", "pctbedft_2011p", 
  "pctwalking", "pctincont_bladr_2011p", "pctincont_bowel_2011p", "pctcath_2011p", "pctchf", "pcthyper", "pctschiz_bipol", 
  "pctvent_2011p", "pctuti", "pctfall30_2011p",  "pctobese", "pctrxdep_2011p", "pctrxpsych_2011p",  "pctrxpsyoff_2011p", "NHCADL_2011p", 
  "NHCpain_2011p", "NHCpu_2011p","obs_rehosprate", "adj_rehosprate", "obs_successfuldc",   
  "obs_medianlos", "mds_hospptyr", "nh_days", "mdshosps"  ,   "avg_dailycensus" ,  "sd_dailycensus") %>% 
  select_if(~!any(is.na(.))) %>%
  na.omit(.)
```

### LASSO  

```{r, eval=F, message=F, warning=F}
library(glmnet)

df_las <- df_cov %>%
  select(-fac_id) %>%
  model.matrix(dc_hosp_any ~ ., data=.) 

cv_fit <- cv.glmnet(df_las[, -1], df_cov$dc_hosp_any, family='binomial', alpha=1, nfold = 10)
```

```{r, eval=F }
plot(cv_fit)
```

### Covariates from LASSO  
```{r, eval=F }
final.vars <- coef(cv_fit, s = "lambda.min")[, 1]

final.vars <- final.vars[final.vars!=0]
final.vars <- final.vars[2:length(final.vars)] %>% #remove intercept  
  sort(.)

final.vars <- names(final.vars)

final.vars
```

mdshosps / mds_hospptyr / fac_hosp_num are all very similar so just lept mds_hospptyr.  

```{r, echo=F}
fac_key_vars <- c('pcthmo', 'agghighcfs', 'obs_rehosprate', 'acuindex2', 'mds_hospptyr', 
                  'fac_ls', 'fac_age_85', 'fac_white', 'fac_black', 'fac_diet', 'fac_dem')
fac_key_lbls <- c('% HMO', 'Agg. High CFS', 'Obs. Rehosp. rate', 'Acuity Index', 'MDS Hosp / pt/ yr', 
                  'No. long-stay', 'LS Hosp. Rate', '% Age 85', '% White', '% Black', '% HMO', '% Mech. Diet', '% Dementia')

cat('Variables selected from 2015 analysis and LASSO: ', fac_key_vars , '\n')
```

```{r, eval=F }
chk_id_vars <- c('age_at_indx', 'age_65', 'age_85', 'female', 
                 'white', 'black', 'hispanic', 'asian', 
                 'm3_adl', 'm3_CFS', 'm3_charlson', 'mds_resp', 
                 'pcthmo', 'agghighcfs', 'obs_rehosprate', 'acuindex2', 'mds_hospptyr', 
                 'fac_ls', 'M3K0510C2', 'mds_dem', 'dc_hosp_any',
                 'M3B0100', 'paymcaid', 'rn2nrs', 'facpoor')
chk_id_lbls <- c('Age at index', 'Age >=65 years', 'Age >=85 years*', 'Female*', 
                 'White', 'Black' ,'Hispanic*', 'Asian*', 
                 'ADL score*', 'CFS score*', 'Charlson score*', 'Resp. diagnosis', 
                 'Facility % HMO', 'Agg. High CFS', 'Obs. Rehosp. rate', ' Acuity index',
                 'MDS hosp. pt yr', 'Facility Long-stay census', 'Mech. Diet', 'Dementia diagnosis', 'Hosp. in season',
                 'Comatose', '% Fac Medicaid', 'RN ratio', 'Poor resource facility')

cat('Key individual or facility variables reported in checking balance: ', chk_id_vars , '\n')
```


```{r, eval=F }
#Only for proportional variables, continuous measures dont have a percentage point diff
pr_vars <- c('female', 'white', 'black', 'hispanic', 'asian', 'mds_resp',
             'mds_dem', 'M3K0510C2', 'age_65', 'age_85')
pr_var_lbls <- c('Female*', 'White', 'Black' ,'Hispanic*', 'Asian*',
                 'Resp. diagnosis', 'Dementia diagnosis', 'Mech. Diet', 
                 'Age >=65 years', 'Age >=85 years*')

cat('Outlier definition, difference in standardized group means:', pr_reject, '\n')
cat('Variables to test for outliers:', paste0(pr_vars, collapse=', '), '\n')
```

## Principal components  

### Scale covariates  
```{r, eval=F }
df_prcomp <- df_cov %>%
  select(-dc_hosp_any) %>%
  distinct(.) 

df_prcomp_scl <- df_prcomp %>%
  select(-fac_id) %>%
  map_dfr(., scale) 

head(df_prcomp_scl)
```

### Compute PC  
```{r, eval=F }
pca_result <- prcomp(df_prcomp_scl, scale = TRUE)  

summary(pca_result)
```

```{r, eval=F }
e_values <- pca_result$sdev[pca_result$sdev>1] 

# First for principal components
df_decomp <- data.frame(pca_result$x) %>%
  bind_cols(., df_prcomp) %>%
  select('fac_id', paste0('PC', 1:length(e_values), sep='')) %>%
  mutate_at(vars(-'fac_id'), scale)


df_pca_weighted <- bind_cols(fac_id=df_decomp$fac_id, 
                             x2=sweep(df_decomp[, 2:21], 2, e_values, "*")) 
```

# Functions for randomization  

```{r, eval=F}
pr_badrand <- function(x, cutoff, n) (sum(abs(x)>=cutoff)/n)*100

do_rand <- function(iter_df) {
   t_diff <- iter_df %>%
      group_by(group) %>%
      summarize_all(mean, na.rm=T) 
    
   #careful refers to global environ
    t_diff <- unlist(t_diff[t_diff$group=='a', chk_id_vars] - t_diff[t_diff$group=='b', chk_id_vars])
    return(t_diff)
}
```

# Method 1. Simple Randomization  

## Assign randomizations  

```{r method 1, warning=F, eval=F}
st_run <- Sys.time()

## Random Assignment - Simple  

df_m1 <- df_cov %>%
  select(fac_id) %>%
  distinct(.) %>%
  pull(.)

#Matrix of values  
m1_res <- list()

m1_res$delta <- matrix(NA, nrow = n_rndms, ncol = length(chk_id_vars)) %>%
  as.data.frame(.)

m1_res$stdev <- df_trial[, chk_id_vars] %>% 
    map_dfr(., sd, na.rm=T) %>%
    t(.)

do_it <- function(x)  {
  sim_iter <- rnd_allot(df_m1) %>%
      as.data.frame(.) %>%
      inner_join(df_trial[, c('fac_id', chk_id_vars)], ., by=c('fac_id'='id')) %>%
      select(-fac_id)
  
  delta <- do_rand(sim_iter)
  return(delta)
}

m1_res$delta[1L:n_rndms, ] <- t(sapply(1L:n_rndms, do_it))
m1_res$smd <- t(apply(m1_res$delta[, 1:length(chk_id_vars)], 1, function(x) x / t(m1_res$stdev)))

st_end <- Sys.time()

cat(paste0(n_rndms), 'Randomizations for M1. Simple Random Assignment \n ')
st_end - st_run
```

## Permute M1  

```{r, message=F, warning=F, eval=F }
st_run <- Sys.time()

library(nloptr)
library(lme4)

m1_rtest <- vector(mode = 'double', length = n_permutes)

for (i  in 1:n_permutes) {
  facs <- rnd_allot(df_m1)

  sim_1 <- df_trial %>%
    select(fac_id, dc_hosp_any) %>%
    inner_join(., facs, by=c('fac_id'='id')) %>%
    mutate(assign = if_else(group=='a', 1L, 0L)) %>%
    na.omit(.)
  
  iter <- glmer(dc_hosp_any ~ assign + (1 | fac_id), data = sim_1, 
                      nAGQ=0, control=glmerControl(optimizer = "nloptwrap"), 
                      family = 'binomial')
  
  m1_rtest[i] <- mean(coef(iter)$fac_id[['assign']])
}


st_end <- Sys.time()
st_end - st_run
```

```{r, eval=F }
save.image(file='flublok_pretrial_sims.RData')
```

# Method 2. Simple stratified randomization - Race, Size  

## Assign Randomizations  

```{r method 2, eval=F }
st_run <- Sys.time()

## 2 Stratum  
  df_m2 <- df_trial %>%
    distinct(fac_id, fac_black, fac_ls) %>%
    mutate(cat_aa = ntile(fac_black, 5),
           cat_fs = ntile(fac_ls, 5),
           strata = interaction(cat_aa, cat_fs)) %>%
    distinct(fac_id, cat_aa, cat_fs, strata)   

#Matrix of values  
m2_res <- list()

m2_res$delta <- matrix(NA, nrow = n_rndms, ncol = length(chk_id_vars))  
m2_res$stdev <- df_trial[, chk_id_vars] %>% 
    map_dfr(., sd, na.rm=T) %>%
    t(.)

do_it <- function(x)  {
  sim_iter <- rnd_str(df_m2, strata, fac_id) %>%
    inner_join(df_trial[, c('fac_id', chk_id_vars)], ., by=c('fac_id')) %>%
    select(group, chk_id_vars)
  
  delta <- do_rand(sim_iter)
  return(delta)
}

m2_res$delta[1:n_rndms, ] <- t(sapply(1L:n_rndms, do_it))
m2_res$smd <- t(apply(m2_res$delta, 1, function(x) x / t(m2_res$stdev)))

st_end <- Sys.time()
cat(paste0(n_rndms), 'Randomizations for M2. Stratified randomization, facility %AA and size quintiles \n ')
st_end - st_run
```

## Permute M2  

```{r, eval=F }
st_run <- Sys.time()

m2_rtest <- vector(mode = 'double', length = n_permutes)

for (i  in 1:n_permutes) {
  
  facs <- rnd_str(df_m2, strata, fac_id)

  sim_1 <- df_trial %>%
    select(fac_id, dc_hosp_any) %>%
    inner_join(., df_m2, by=c('fac_id'='fac_id')) %>%
    inner_join(., facs, by=c('fac_id'='fac_id')) %>%
    mutate(assign = if_else(group=='a', 1L, 0L)) %>%
    na.omit(.)
  
  iter <- lme4::glmer(dc_hosp_any ~ assign + factor(cat_aa) + factor(cat_fs) + (1 | fac_id), 
              nAGQ=0, control=glmerControl(optimizer = "nloptwrap"), 
              data = sim_1, family = 'binomial')
  
  m2_rtest[i] <- mean(coef(iter)$fac_id[['assign']])
}


st_end <- Sys.time()
st_end - st_run
```

```{r, eval=F }
save.image(file='flublok_pretrial_sims.RData')
```

# Method 3. Pair-matched Randomization - Mahalanobis Distance  

## Create Pair-Matches  

```{r, warning=F, message=F, eval=F}
library(nbpMatching)

# create a covariate matrix
  df_mtch <- df_trial %>%
    select(fac_id, fac_key_vars) %>%
    as.data.frame(.) %>%
    distinct(.)

# create distances
  df_dist <- gendistance(df_mtch, idcol=1)

# create distancematrix object
  df_mdm <- distancematrix(df_dist)

# create matches
  df_mtch_2 <- nonbimatch(df_mdm)

# review quality of matches
  df_qom <- qom(df_dist$cov, df_mtch_2$matches)

  df_mtch_3 <- df_mtch_2$matches %>%
    select(Group1.ID, Group2.ID) %>%
    mutate(class = row_number()) %>%
    tidyr::pivot_longer(, cols = c('Group1.ID', 'Group2.ID'),
                        values_to = 'fac_id') %>%
    select(fac_id, class) %>%
    arrange(fac_id) %>%
    group_by(fac_id) %>%
      slice(1) %>%
    ungroup(.)
    
## Mahalanobis distane
  df_m3 <- df_trial %>%
    select(fac_id) %>%
    inner_join(., df_mtch_3) %>%
    distinct(.)

  n_distinct(df_mtch_3$fac_id)
```

```{r method 3, eval=F }
st_run <- Sys.time()

#Matrix of values  
m3_res <- list()

m3_res$delta <- matrix(NA, nrow = n_rndms, ncol = length(chk_id_vars))  
m3_res$stdev <- df_trial[, chk_id_vars] %>% 
    map_dfr(., sd, na.rm=T) %>%
    t(.)

do_it <- function(x)  {
  sim_iter <- rnd_str(df_m3, class, fac_id) %>%
    inner_join(df_trial[, c('fac_id', chk_id_vars)], ., by=c('fac_id')) %>%
    select(group, chk_id_vars)
  
  delta <- do_rand(sim_iter)
  return(delta)
}

m3_res$delta[1:n_rndms, ] <- t(sapply(1L:n_rndms, do_it))
m3_res$smd <- t(apply(m3_res$delta, 1, function(x) x / t(m3_res$stdev)))

st_end <- Sys.time()
cat(paste0(n_rndms), 'Randomizations for M3. Pair-Matched randomization, Key Variables \n ')
st_end - st_run
```

## Permute M3  

```{r, eval=F }
st_run <- Sys.time()

m3_rtest <- vector(mode = 'double', length = n_permutes)

for (i  in 1:n_permutes) {
  
  facs <- rnd_str(df_m3, class, fac_id)

  sim_1 <- df_trial %>%
    select(fac_id, dc_hosp_any) %>%
    inner_join(., facs, by=c('fac_id'='fac_id')) %>%
    mutate(assign = if_else(group=='a', 1L, 0L)) %>%
    na.omit(.)
  
  iter <- lme4::glmer(dc_hosp_any ~ assign + (1 | class) + (1 | fac_id), 
                      nAGQ=0, control=glmerControl(optimizer = "nloptwrap"), 
                      data = sim_1, family = 'binomial')
  
  m3_rtest[i] <- mean(coef(iter)$fac_id[['assign']])
}


st_end <- Sys.time()
st_end - st_run
```

```{r, eval=F }
save.image(file='flublok_pretrial_sims.RData')
```

# Method 4. K-means clustering, key variables  

```{r method 4, eval=F }
st_run <- Sys.time()

#Matrix of values  
df_km_1 <- df_trial %>%
  select(fac_id, fac_key_vars) %>%
  distinct(.) %>%
  mutate_at(vars(fac_key_vars), scale)

repeat {
  df_km_2 <- df_km_1 %>%
    select(-fac_id) %>%
    as.matrix(.) %>%
    kmeans(x=., centers = k_clust, nstart = k_starts, iter.max=k_iters)
  
  df_m4 <- bind_cols(fac_id  = df_km_1$fac_id,
                         strata = df_km_2$cluster) %>%
    distinct(.)
    
    if (min(table(df_m4$strata)) <= 2) {
      break
    }
  k_clust <- k_clust + 1L
}

#Matrix of values  
m4_res <- list()

m4_res$delta <- matrix(NA, nrow = n_rndms, ncol = length(chk_id_vars))  
m4_res$stdev <- df_trial[, chk_id_vars] %>% 
    map_dfr(., sd, na.rm=T) %>%
    t(.)

do_it <- function(x)  {

  sim_iter <- rnd_str(df_m4, strata, fac_id) %>%
    inner_join(df_trial[, c('fac_id', chk_id_vars)], ., by=c('fac_id')) %>%
    select(group, chk_id_vars)
  
  delta <- do_rand(sim_iter)
  return(delta)
}

m4_res$delta[1:n_rndms, ] <- t(sapply(1:n_rndms, do_it))
m4_res$smd <-t(apply(m4_res$delta, 1, function(x) x / t(m4_res$stdev)))

st_end <- Sys.time()

cat(paste0(n_rndms), 'Method 4. K-means clustering on key variables \n')
st_end - st_run
```

## Permute M4  

```{r, eval=F }
st_run <- Sys.time()

m4_rtest <- vector(mode = 'double', length = n_permutes)

for (i  in 1:n_permutes) {
  
  facs <- rnd_str(df_m4, strata, fac_id)

  sim_1 <- df_trial %>%
    select(fac_id, dc_hosp_any) %>%
    inner_join(., facs, by=c('fac_id'='fac_id')) %>%
    mutate(assign = if_else(group=='a', 1L, 0L)) %>%
    na.omit(.)
  
  iter <- lme4::glmer(dc_hosp_any ~ assign + (1 | strata) + (1 | fac_id), 
                      nAGQ=0, control=glmerControl(optimizer = "nloptwrap"), 
                      data = sim_1, family = 'binomial')
  
  m4_rtest[i] <- mean(coef(iter)$fac_id[['assign']])
}


st_end <- Sys.time()
st_end - st_run
```

```{r }
save.image(file='flublok_pretrial_sims.RData')
```

# Method 5. K-means, PCA cluster eigen-weighted  

```{r method 5, eval=F }
st_run <- Sys.time()

#Matrix of values  
df_pca_1 <- df_pca_weighted 

repeat {
  df_km_2 <- df_pca_1 %>%
    select(-fac_id) %>%
    as.matrix(.) %>%
    kmeans(x=., centers = k_clust, nstart = k_starts, iter.max=k_iters)
  
  df_m5 <- bind_cols(fac_id  = df_pca_1$fac_id,
                         strata = df_km_2$cluster) %>%
    distinct(.)
    
    if (min(table(df_m5$strata)) <= 2) {
      break
    }
  k_clust <- k_clust + 1L
}

#Matrix of values  
m5_res <- list()

m5_res$delta <- matrix(NA, nrow = n_rndms, ncol = length(chk_id_vars))  
m5_res$stdev <- df_trial[, chk_id_vars] %>% 
    map_dfr(., sd, na.rm=T) %>%
    t(.)

do_it <- function(x)  {

  sim_iter <- rnd_str(df_m5, strata, fac_id) %>%
    inner_join(df_trial[, c('fac_id', chk_id_vars)], ., by=c('fac_id')) %>%
    select(group, chk_id_vars)
  
  delta <- do_rand(sim_iter)
  return(delta)
}

m5_res$delta[1:n_rndms, ] <- t(sapply(1:n_rndms, do_it))
m5_res$smd <-t(apply(m5_res$delta, 1, function(x) x / t(m5_res$stdev)))

st_end <- Sys.time()

cat(paste0(n_rndms), 'Method 5. K-means on PCA \n')
st_end - st_run
```

## Permute M5  

```{r, eval=F }
st_run <- Sys.time()

m5_rtest <- vector(mode = 'double', length = n_permutes)

for (i  in 1:n_permutes) {
  
  facs <- rnd_str(df_m5, strata, fac_id)

  sim_1 <- df_trial %>%
    select(fac_id, dc_hosp_any) %>%
    inner_join(., facs, by=c('fac_id'='fac_id')) %>%
    mutate(assign = if_else(group=='a', 1L, 0L)) %>%
    na.omit(.)
  
  iter <- lme4::glmer(dc_hosp_any ~ assign + (1 | strata) + (1 | fac_id), 
                      nAGQ=0, control=glmerControl(optimizer = "nloptwrap"), 
                      data = sim_1, family = 'binomial')
  
  m5_rtest[i] <- mean(coef(iter)$fac_id[['assign']])
}


st_end <- Sys.time()
st_end - st_run
```

```{r, eval=F }
save.image(file='flublok_pretrial_sims.RData')
```

# Method 6. Re-randomization  

## Find Acceptable Cut-Off Value  
```{r method 6, eval=F }
st_run <- Sys.time()

  df_rerand <- df_cov %>%
    select(-dc_hosp_any) %>%
    distinct(.) %>%
    mutate_at(vars(-fac_id), scale)

 # Find M-distance cutoff empirically   
  do_it <- function(x) {

    sim_iter <- rnd_allot(df_fac) %>%
      as.data.frame(.) %>%
      inner_join(df_rerand, ., by=c('fac_id'='id')) %>%
      select(-fac_id) %>%
      as.data.frame(.)

    mdis_iter <- mahalanobis(colMeans(sim_iter[sim_iter$group=='a', fac_key_vars], na.rm=T), 
                               colMeans(sim_iter[sim_iter$group=='b', fac_key_vars], na.rm=T), 
                                  cov= cov(sim_iter[, fac_key_vars])) * 
                   ((nrow(sim_iter[sim_iter$group=='a', ]) * nrow(sim_iter[sim_iter$group=='b', ])) / (nrow(sim_iter)))
  }

  m6_chk_mdis <- sapply(1:n_rndms, do_it)
  
  m6_cutoff <- quantile(m6_chk_mdis, 0.001)  
  
st_end <- Sys.time()

cat('Method 6. Re-randomization cut-off', round(m6_cutoff, 3), ' \n')
st_end - st_run
```

## Perform Re-randomizations  

```{r, eval=F }
st_run <- Sys.time()

  m6_res <- list()
  
  m6_res$delta <- matrix(NA, nrow = n_re_rndms, ncol = length(chk_id_vars))  
  m6_res$stdev <- df_trial[, chk_id_vars] %>% 
      map_dfr(., sd, na.rm=T) %>%
      t(.)

  df_m6 <- df_trial %>%
    select(fac_id, fac_key_vars) %>%
    distinct()   
  
  do_it <- function(x)  {
    repeat { # search for good rerand
    
      chk_rand <- rnd_allot(df_fac) %>%
        inner_join(df_m6[, c('fac_id', fac_key_vars)], ., by=c('fac_id'='id')) 
    
      mdis_iter <- Rfast::mahala(colMeans(chk_rand[chk_rand$group=='a', fac_key_vars], na.rm=T), 
                                 colMeans(chk_rand[chk_rand$group=='b', fac_key_vars], na.rm=T), 
                                 sigma= cov(chk_rand[, fac_key_vars])) * 
                       ((nrow(chk_rand[chk_rand$group=='a', ]) * nrow(chk_rand[chk_rand$group=='b', ])) / (nrow(chk_rand)))
      
      if(mdis_iter<m6_cutoff) {
        break
      }
    }
  
  sim_iter <- inner_join(df_trial[, c('fac_id', chk_id_vars)], 
                         chk_rand[, c('fac_id', 'group')], 
                         by=c('fac_id')) %>%
    select(group, chk_id_vars)
  
  delta <- do_rand(sim_iter)
  return(delta)
}

m6_res$delta[1:n_rndms, ] <- t(sapply(1L:n_rndms, do_it))
m6_res$smd <- t(apply(m6_res$delta, 1, function(x) x / t(m6_res$stdev)))

st_end <- Sys.time()

cat(paste0(n_rndms), 'Method 6. Re-randomizations, \n')
st_end - st_run
```

## Permute M6  

```{r, eval=F }
st_run <- Sys.time()

m6_rtest <- vector(mode = 'double', length = n_permutes)

for (i  in 1:n_permutes) {
  
      repeat { # search for good rerand
    
      chk_rand <- rnd_allot(df_fac) %>%
        inner_join(df_m6[, c('fac_id', fac_key_vars)], ., by=c('fac_id'='id')) 
    
      mdis_iter <- Rfast::mahala(colMeans(chk_rand[chk_rand$group=='a', fac_key_vars], na.rm=T), 
                                 colMeans(chk_rand[chk_rand$group=='b', fac_key_vars], na.rm=T), 
                                 sigma= cov(chk_rand[, fac_key_vars])) * 
                       ((nrow(chk_rand[chk_rand$group=='a', ]) * nrow(chk_rand[chk_rand$group=='b', ])) / (nrow(chk_rand)))
      
      if(mdis_iter<m6_cutoff) {
        break
      }
    }
  
  sim_1 <- df_trial %>%
    select(fac_id, dc_hosp_any) %>%
    inner_join(., chk_rand[, c('fac_id', 'group')], by=c('fac_id')) %>%
    mutate(assign = if_else(group=='a', 1L, 0L)) %>%
    na.omit(.)
  
  
  iter <- lme4::glmer(dc_hosp_any ~ assign + (1 | fac_id), 
                      nAGQ=0, control=glmerControl(optimizer = "nloptwrap"), 
                      data = sim_1, family = 'binomial')
  
  m6_rtest[i] <- mean(coef(iter)$fac_id[['assign']])
}


st_end <- Sys.time()
st_end - st_run
```

```{r, eval=F }
save.image(file='flublok_pretrial_sims.RData')
```

# Evaluation of Methods  

## Distribution of Variables  

## Gather checking variables from Methods  

```{r, eval=F }
fct_lvls <- rndm_methods

for (i in 1:6) { #loop to create smd objects
  assign(paste0('m', i, '_diff'), get(paste0('m', i, '_res'))$delta %>%
     as.data.frame(.) %>%
    mutate(Method = fct_lvls[i]))
}

diff_all <- mget(ls()[str_detect(ls(), '_diff')]) #get all objects with '_diff' names

diff_all <- bind_rows(diff_all) %>%
  mutate(Method = factor(Method, levels= fct_lvls)) %>%
  na.omit(.)

colnames(diff_all) <- c(chk_id_vars, 'Method')

diff_all_grp <- gather(diff_all, key = 'var', value = 'x', -Method) %>%
  mutate(var = factor(var, levels=chk_id_vars, labels = chk_id_lbls))
```

## Compute range of values  

```{r, eval=F }
cats <- c(0.025, 0.25, 0.5, 0.75, 0.975)

quant_vals <- function(x) {
  y <- quantile(x, probs=cats)
  return(y)
}
```

```{r, message=F, warning=F, eval=F}

cats_2 <- c(rep(cats[1], 6*length(chk_id_vars)), 
          rep(cats[2], 6*length(chk_id_vars)), 
          rep(cats[3], 6*length(chk_id_vars)), 
          rep(cats[4], 6*length(chk_id_vars)), 
          rep(cats[5], 6*length(chk_id_vars)))

sum_diff_grp <- diff_all_grp %>%
  split(., list(.$Method, .$var)) %>%
  purrr::map(~quant_vals(.$x)) %>%
  bind_rows(.) %>%
  pivot_longer(., 
               cols = everything(),
               values_to = 'x') %>%
  mutate(Quintile = cats_2) %>%
  separate(., name, into = c('Method', 'var'), sep = "([\\.])") 
```

### Table of Quintiles  

```{r, eval=F }

tab_quint <- sum_diff_grp %>%
  pivot_wider(., id_cols = c(Method, var), names_from = Quintile, values_from = x) %>%
  select(var, everything()) 

tab_quint[, 3:7] <- round(tab_quint[, 3:7], 3) 

DT::datatable(tab_quint)
```

### Variables greater 0.2 SMD  

```{r, eval=F }
fct_lvls <- rndm_methods

for (i in 1:6) { #loop to create smd objects
  assign(paste0('m', i, '_smd'), get(paste0('m', i, '_res'))$smd %>%
     as.data.frame(.) %>%
    mutate(Method = fct_lvls[i]))
}

smd_all <- mget(ls()[str_detect(ls(), '_smd')]) #get all objects with '_smd' names

smd_all <- bind_rows(smd_all) %>%
  mutate(Method = factor(Method, levels= fct_lvls)) %>%
  na.omit(.)

colnames(smd_all) <- c(chk_id_vars, 'Method')

smd_all_grp <- gather(smd_all, key = 'var', value = 'x', -Method) %>%
  mutate(var = factor(var, levels=chk_id_vars, labels = chk_id_lbls))
```

#### Table % of variables with >0.2 SMD  

```{r, eval=F }
prop_outlier <- smd_all_grp %>%
  mutate(out = if_else(abs(x)>0.2, 1L, 0L)) %>%
  group_by(Method, var) %>%
  summarize(pr_out = (sum(out) / n())*100) %>%
  pivot_wider(., id_cols = c(var), names_from = Method, values_from = pr_out)

DT::datatable(prop_outlier)
```

### Variance reduction  

```{r, eval=F }
var_all_grp <- diff_all_grp %>%
  group_by(Method, var) %>%
  summarize(variance = var(x)) %>%
  pivot_wider(., id_cols = c(var), names_from = Method, values_from = variance)

for (i in 3:7) {
  var_all_grp[, i] <- (var_all_grp[, i] - var_all_grp[, 2]) / var_all_grp[, 2]
}

prvred <- var_all_grp[, -2] 

prvred[nrow(prvred)+1, 2:6] <- colMeans(prvred[, 2:6])
prvred[nrow(prvred)+1, 2:6] <- sapply(prvred[, 2:6], min)
prvred[nrow(prvred)+1, 2:6] <- sapply(prvred[, 2:6], max)  

DT::datatable(prvred) %>%
  DT::formatRound(c(2:6), 3)
```

### Permutation Results  

```{r, eval=F }
for (i in 1:6) { #loop to create smd objects
  assign(paste0('m', i, '_diff'), get(paste0('m', i, '_res'))$delta %>%
     as.data.frame(.) %>%
    mutate(Method = fct_lvls[i]))
}

rtest_all <- mget(ls()[str_detect(ls(), '_rtest')]) #get all objects with '_smd' names

rtest_all <- bind_rows(rtest_all)
colnames(rtest_all) <- fct_lvls 

rtest_all_grp <- pivot_longer(rtest_all, cols = everything(), names_to = 'Method', values_to = 'x') %>%
  arrange(Method) %>%
  group_by(Method) %>%
  summarize(lci = quantile(x, probs = c(0.025)),
            uci = quantile(x, probs = c(0.975)))  

DT::datatable(rtest_all_grp) %>%
  DT::formatRound(c(2:6), 4)
```

```{r, eval=F }
save.image(file='flublok_pretrial_sims.RData')
```
